Type-Safe Modular Parsing

### 2017-01-14 paper/sections/Inheritance.tex
small typo
Typos in S1234.

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO
style~\cite{wadler1998expression}. On the other hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators that also generate
AST code (such as ANTLR~\cite{antlr1995}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

@@

It is actually quite simple to let \inlinecode{pExpr} cover the newly
extended case without modifying existing code. Method overriding is a
standard feature which often comes with inheritance, and it allows us
to redefined an inherited method, such as \inlinecode{pExpr}. We can
build the new parser which correctly parses \inlinecode{"1 + x"}
through overriding:

@@

The \inlinecode{super[T].x} syntax in Scala, so called \textit{static super reference}, refers to the type or method \inlinecode{x} in the parent trait \inlinecode{T}. Under multiple inheritance, it can be used to distinguish the method of the same name. Therefore in the new parser, we use \inlinecode{super} to specify which \inlinecode{pExpr} we are referring to.

@@

\paragraph{Conflicts and/or ambiguity} As we demonstrated before, the longest match alternative combinator is useful when composing parsers. However, ambiguity grammars still cause subtle issues in modular parsing. For example, if \inlinecode{parserA} and \inlinecode{parserB} accept a same string but generate different results, then the result of \inlinecode{parserA ||| parserB} is really depends on the implementation of the alternative combinator, because the results can not be prioritized by length. Since the composition is handled explicitly by user, conflicts could be resolved by overriding ambiguous parsing code, using the pattern discussed in Section~\ref{subsec:overriding-rules}.

>>>>

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO
style~\cite{wadler1998expression}. On the other hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators also generate
AST code (such as ANTLR~\cite{antlr1995}), normally the generated code is based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

@@

It is actually quite simple to let \inlinecode{pExpr} cover the newly
extended case without modifying existing code. Method overriding is a
standard feature which often comes with inheritance, and it allows us
to redefine an inherited method, such as \inlinecode{pExpr}. We can
build the new parser which correctly parses \inlinecode{"1 + x"}
through overriding:

@@

The \inlinecode{super[T].x} syntax in Scala, so called \textit{static super reference}, refers to the type or method \inlinecode{x} in the parent trait \inlinecode{T}. Under multiple inheritance, it can be used to distinguish the methods of the same name. Therefore in the new parser, we use \inlinecode{super} to specify which \inlinecode{pExpr} we are referring to.

@@

\paragraph{Conflicts and/or ambiguity} As we demonstrated before, the longest match alternative combinator is useful when composing parsers. However, ambiguous grammars still cause subtle issues in modular parsing. For example, if \inlinecode{parserA} and \inlinecode{parserB} accept a same string but generate different results, then the result of \inlinecode{parserA ||| parserB} really depends on the implementation of the alternative combinator, because the results can not be prioritized by length. Since the composition is handled explicitly by user, conflicts could be resolved by overriding ambiguous parsing code, using the pattern discussed in Section~\ref{subsec:overriding-rules}.

### 2017-01-14 paper/sections/ObjectAlgebras.tex
Typos in S1234.

The inheritance-based approach allows building extensible parsers, based on an OO class hierarchy.  Nevertheless, the addition of new operations over ASTs is problematic using traditional OO ASTs. In this section, we
show how to support both the forms of extensibility for ASTs (easy
addition of language constructs, and easy addition of operations)
using Object Algebras~\cite{Oliveira:2012}.

@@

\noindent Then all classes represented language constructs could be 
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work 
by Wang and Oliveira~\cite{wang2016expression}, which show a technique that is type-safe
in many cases). More importantly, a second problem is that
\begin{comment}
But then programmers have to define a new class for each construct by
extending the old class with \lstinline{FVarsExpr}, and implementing
the corresponding \lstinline{fVars} method:

@@

Besides introducing boilerplate code, more classes have to be defined
along with future extensions. They pollutes the namespace heavily.
More importantly, 
\end{comment}
is that even if that approach would work, the parsing code would not
be reusable. \lstinline{VarExprParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser has type \inlinecode{Parser[Expr]}, which indicates the result have type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

@@

Object Algebras captures a design pattern that address the Expression
Problem, achieving two dimensions of extensibility (language
constructs and operations) in a modular and type-safe way. The
definition of data structures is separated from their behaviours, and
future extensions on both dimensions no longer require existing code
to be modified, supporting separate compilation.

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty print the expression,
we pass an instance of \inlinecode{VarExprPrint} to \inlinecode{makeExp}:

@@

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method override for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\lstinputlisting[linerange=23-30]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use fields defined by keyword \inlinecode{def}, can not use those defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

@@

\noindent The type of the object algebra field \lstinline{alg} is first refined
to \lstinline{VarExpAlg[E]}, to allow calling the additional factory method
for variables. Note that, unlike the previous attempt, \lstinline{VarExpAlg[E]}
is a subtype of \lstinline{ExpAlg[E]}, thus the type-refinement of the
field is allowed. Now, the code for parsing variables (\lstinline{pVar}) can
call \lstinline{alg.varE}. The following code illustrates how to use
the parser from a client perspective:

@@

In the client code shown above, we first need to choose a concrete
object algebra to initialize the \lstinline{alg} field in \lstinline{VarExprOAParser}. Here we pick the
pretty-printing algebra \lstinline{VarExprPrint}, but any other object
algebra that implements \lstinline{VarExprAlg} would work.
With an instance of \lstinline{VarExprOAParser} in hand, we can call
\lstinline{pE} to obtain the parser to feed to the \lstinline{parse} method.
The result shows that such a pattern provides modular parsing.

Note that, similarly to the approach in Section~\ref{sec:inheritance} independent extensibility is also supported via multiple trait inheritance.
Since independent extensibility is achieved using essentially the same technique as in Section~\ref{sec:inheritance} we omit that code here.

>>>>

The inheritance-based approach allows building extensible parsers, based on an OO class hierarchy.  Nevertheless, the addition of new operations over ASTs is problematic using traditional OO ASTs. In this section, we
show how to support both forms of extensibility on ASTs (easy
addition of language constructs, and easy addition of operations)
using Object Algebras~\cite{Oliveira:2012}.

@@

\noindent Then all classes representing language constructs could be 
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work 
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
\begin{comment}
But then programmers have to define a new class for each construct by
extending the old class with \lstinline{FVarsExpr}, and implementing
the corresponding \lstinline{fVars} method:

@@

Besides introducing boilerplate code, more classes have to be defined
along with future extensions. They pollutes the namespace heavily.
More importantly, 
\end{comment}
is that even if that approach would work, the parsing code would not
be reusable. \lstinline{VarExprParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser has type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

@@

Object Algebras capture a design pattern that addresses the Expression
Problem, achieving two dimensions of extensibility (language
constructs and operations) in a modular and type-safe way. The
definition of data structures is separated from their behaviours, and
future extensions on both dimensions no longer require existing code
to be modified, supporting separate compilation.

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression,
we pass an instance of \inlinecode{VarExprPrint} to \inlinecode{makeExp}:

@@

\lstinputlisting[linerange=8-16]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\lstinputlisting[linerange=23-30]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use methods defined by keyword \inlinecode{def}, but cannot access fields defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

@@

\noindent The type of the Object Algebra field \lstinline{alg} is first refined
to \lstinline{VarExpAlg[E]}, to allow calling the additional factory method
for variables. Note that, unlike the previous attempt, \lstinline{VarExpAlg[E]}
is a subtype of \lstinline{ExpAlg[E]}, thus the type-refinement of the
field is allowed. Now, the code for parsing variables (\lstinline{pVar}) can
call \lstinline{alg.varE}. The following code illustrates how to use
the parser from a client's perspective:

@@

In the client code shown above, we first need to choose a concrete
Object Algebra to initialize the \lstinline{alg} field in \lstinline{VarExprOAParser}. Here we pick the
pretty-printing algebra \lstinline{VarExprPrint}, but any other Object
Algebra that implements \lstinline{VarExprAlg} would work.
With an instance of \lstinline{VarExprOAParser} in hand, we can call
\lstinline{pE} to obtain the parser to feed to the \lstinline{parse} method.
The result shows that such a pattern provides modular parsing.

Note that, similar to the approach in Section~\ref{sec:inheritance} independent extensibility is also supported via multiple trait inheritance.
Since independent extensibility is achieved using essentially the same technique as in Section~\ref{sec:inheritance} we omit that code here.

### 2017-01-14 paper/sections/PackratParsing.tex
Typos in S1234.

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The top-down, recursive descent parsing strategy adopted by those parser combinator libraries cannot support left-recursive grammars directly. The common solution is to rewrite the grammar into an equivalent but not left-recursive one, so called left-recursion elimination.

@@

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the information of left associativity is lost. The parse tree
must be transformed to recover the correct syntactic structure.

@@

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longest match alternative combinator.
%Below we present an example to illustrate Scala parsers.

>>>>

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to compose them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and so are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The top-down, recursive descent parsing strategy adopted by those parser combinator libraries cannot support left-recursive grammars directly. The common solution is to rewrite the grammar into an equivalent but non-left-recursive one, called left-recursion elimination.

@@

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the information of left-associativity is lost. The parse tree
must be transformed to recover the correct syntactic structure.

@@

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longest match alternative combinator.
%Below we present an example to illustrate Scala parsers.
